EuPPollNet: A European database of plant-pollinator networks

### 2024-05-29 Manuscript/EuPPollNet_word.qmd
Minor typo edits

The interaction between plants and pollinators is one of the most well-documented mutualisms on Earth. Plant-pollinator interactions involve a great diversity of species, largely attributed to their coevolutive history [@ollerton2017], and are critically important for terrestrial biodiversity and the human economy. The synergistic effects of climate change with other global change pressures are threatening worldwide biodiversity [@sala2000; @bellard2014], including plant and pollinator species as well as their interactions [@goulson2015; @settele2016; @eichenberg2021]. Under this scenario, the increasing availability of biodiversity data plays a major role in our ecological understanding of species and guiding conservation planning [@heberling2021]. However, our knowledge of plant and pollinator species and their network of interactions still exhibits major temporal, spatial and taxonomic biases [@archer2014; @troia2016; @poisot2021; @marshall2024], limiting our ability to effectively protect their biodiversity.

@@

Here, we present the EuPPollNet database, an acronym derived from European plant-pollinator networks, which contains harmonized information on interaction data of plants and pollinators at European level. The primary focus of the pollinator taxonomic groups is on the main orders of entomofauna that visit and pollinate flowering plants in Europe. These include insect species from the orders Hymenoptera, Diptera, Lepidoptera, and Coleoptera [@potts2015], which comprise almost the totality of recorded interactions in the EuPPollNet database (`r format(round(main_orders_percent, 2), nsmall = 2)`%). To understand the scope of the database, we examine the taxonomic and sampling coverage of the different plant and pollinator species at European level with the help of the most up-to-date species checklists and rarefaction analyses. In addition, we investigate how two key structural metrics of plant-pollinator networks, such as connectance and nestedness, change across the latitudinal range of studies and bioclimatic regions. Finally, we also examine whether plant-pollinator networks fulfill the expectation of being more nested than null expectations. EuPPollNet aims to cover a wide range of taxonomic groups and habitats, while also providing other variables of interest that allow to better control for the ecological context and sampling methods. In addition, EuPPollNet offers a transparent and accessible workflow of its data management and species harmonization that allow to being reused and keep building on it over time. We expect that this database will help to evaluate macro-ecological processes and current gaps of plant-pollinator interactions at European level.

@@

For plants: (i) we initially verified the exact matches against the GBIF species checklist; (ii) selected unmatched cases and fix orthographic errors; (iii) retrieved again taxonomic information for those unmatched cases, evaluated accuracy of fuzzing matching and programmatically fixed records that were still not found; (iv) finally, we used the World Flora Taxonomic Backbone [@govaerts2021\; WFO, July 7, 2022] as the ultimate filter for taxonomic information as we used it to calculate the plant taxonomic coverage of our database.

For pollinators: (i) we first created a checklist of species names for the most representative pollinator groups at European level by combining the recently published checklists of bees [@ghisbain2023], syrphids [@kovcic2023] and butterflies [@wiemers2018]; (ii) then, we compared pollinator species names against the checklist and recovered some unmatched cases with restrictive fuzzy matching by using **stringdist** package [@van2014]; (iii) we programmatically fixed unmatched records when necessary and retrieved the taxonomic information for all species from GBIF; (iv) we fixed the non-found cases in the GBIF checklist and made sure that all species names from bees, syrphids and butterflies were named according to their respective species checklists.

@@

To assess the completeness of plant and pollinator species in the EuPPollNet database at European level, we used the aforementioned checklists for plants and pollinators. Specifically for plants, we refined the checklist to include only European flowering plants and excluded taxonomic groups not associated with biotic pollination. We did this by first excluding the families considered to have exclusively a wind pollination mode [see @culley2002], and then by filtering out the genera with wind or non-biotic pollination from families that exhibit both biotic and non-biotic pollination modes. Additionally, we manually included some exotic species and added unresolved species names that were not present in the accepted names of the checklist at the current version of usage. For pollinators, we compared only the taxonomic coverage of bees, syrphids and butterflies by using their species checklists at European level. Finally, to evaluate if the presence-absence of interaction records for bees and flowering plants follows a phylogenetic pattern within the database, we calculated with the the phylogenetic signal with the help of the function _phylosig_ from the **phytools** package.

@@

The estimated sampling coverage of plant and pollinator species within the database across the different sampling sites is approximately `r round(poll_coverage)`% for both taxonomic groups. This suggest that the rarefied accumulation curves of both plant and pollinator species exhibit already a "quasi-asymptotic" growth of species richness by considering the current number of sampling sites or networks (**Figure 2a-2b**). The predicted observed species richness by doubling the sampling effort on the already sampled habitat types within the database will only increase pollinator richness by `r format(round(poll_increase3, 2), nsmall=2)`% and plant richness by `r format(round(plant_increase3, 2), nsmall=2)`%. However, the sampling coverage of interactions is `r format(round(int_coverage, 2), nsmall=2)`%, and by doubling the sampling effort the predicted number of unique interactions recorded will have approximately a twofold increase (`r format(round(int_increase3, 2), nsmall=2)`%; **Figure 2c**). Similarly, to the plant and pollinator species accumulation curves across sampling sites, when we consider the accumulated pollinator richness across sampled plant species, this curve also shows a "quasi-asymptotic" growth with a sampling coverage value of `r format(round(plant_poll_coverage, 2), nsmall=2)`%. The predicted recorded pollinator species by doubling the number of plants sampled is expected to increase by `r format(round(plant_poll_increase3, 2), nsmall=2)`% (**Figure 2d**). In addition, when we explore how common or rare are plant and pollinator species across all sampling sites, we find that the majority of plant and pollinator species tend to be regionally specific, and only a minor portion of them is being shared across a broad range of sampling sites (**Figures 2e-2f**). For instance, most plants (`r format(round(plants_under_one_percent, 2), nsmall=2)`%) and pollinators (`r format(round(polls_under_one_percent, 2), nsmall=2)`%) are exclusively found in less than 1% of sampling sites; and the most common plant (*`r most_common_plant`*) and pollinator (*`r most_common_poll`*) species are only found in `r format(round(most_common_plant_percent, 2), nsmall=2)`% and `r format(round(most_common_poll_percent, 2), nsmall=2)`% of sampling sites, respectively.

```{r echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE, out.width="95%", fig.cap= "\\textbf{Figure 2}. Graphs (a-b-c) indicate the accumulation curves for pollinators, plants, and their interactions across sampling sites. Grey solid lines represent 100 randomized accumulation curves, the black solid lines represents the interpolated curve (i.e., the mean across curves), and the red dashed lines illustrates the extrapolated curve for approximately 3000 sampling sites. The solid black points indicate the number of species and interactions contained in the database. Graph (d) shows the accumulation curve of pollinator species across increasing number of plant species. This last graph uses the same color and shape structure as the ones in the top panel. Graphs (e-f) indicate the percentage of occurrence (i.e., incidence) of plant and pollinator species across sampling sites. Species on the left (i.e., common) are found in many sampling sites, while species on the right (i.e., rare) are found in few or only a single sampling site.", fig.width=10, fig.height=6, out.width="115%"}

@@

Connectance and nestedness did not change across the latitudinal range of studies (connectance: R$^2$ = `r round(connectance_lat_r2, 2)`, _P_ = `r round(connectance_lat_pval, 2)`, **Figure 6a**; NODFc: R$^2$ `r round(normalised_nestedness_lat_r2, 2)`, _P_ = `r round(normalised_nestedness_lat_pval, 2)`, **Figure 6b**). However, differences were found for connectance but not nestedness across bioclimatic regions. Specifically, the alpine region differed statistically in connectance to the other bioclimatic regions (Tukey test _P_ < 0.01). In addition, the atlantic and continental bioclimatic regions also showed statistical meaningful differences in connectance (Tukey test _P_ < 0.01). Connectance values ranged between `r round(min_connectance, 2)` to `r round(max_connectance, 2)` ($\bar{x}$ = `r round(mean_connectance, 2)`) and followed an negative exponential relationship with the geometric mean of plant and pollinator species per network (Kendall $\tau$ = `r round(connectance_spp_mean_corr_tau, 2)`, _P_ < 0.01; **Figure 6c**). Nestedness values (NODFc) ranged between `r round(min_nodfc, 2)` to `r round(max_nodfc, 2)` ($\bar{x}$ = `r round(mean_nodfc, 2)`), and were independent of the mean number of species (Kendall $\tau$ = `r round(normalised_nestedness_spp_mean_corr_tau, 2)`, _P_ = `r round(normalised_nestedness_spp_mean_corr_pval, 2)`; **Figure S3**). Finally, we found that `r round(over_z_percent, 2)`% of networks were statistically more nested than expected by chance, `r round(no_diff_z_percent, 2)`% showed no statistical difference, and `r round(under_z_percent, 2)`% were less nested than null expectations (**Figure 6d**).

```{r echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE, out.width="95%", fig.cap= "\\textbf{Figure 6}. Graphs (a-b) indicate network connectance and nestedness across the latitudinal range of the studies from the database. In addition, the bioclimatic region where each network is located is indicated with points of different shapes and colours. Graph (c) shows the the association between network connectances and the geometric mean of plant and pollinator species per network. Graph (d) shows the distribution of z-scores when comparing the nestedness from the empirical networks with their randomised counterparts (100 null models for each network). The vertical red dashed lines represent the z critical value for a two tailed test with alpha = 0.05. Z-scores to the left of the first vertical red dashed line indicate that networks are less nested than expected by chance (red), those between the two dashed lines indicate no statistical difference from random expectations (green), and those to the right indicate that networks are more nested than expected by chance (blue).", fig.width=10, fig.height=7, fig.align='center'}

# Load libraries

library(psych)
# Load data
data = readRDS("../Data/3_Final_data/Interaction_data.rds")
metrics_by_network = readRDS("../Data/Working_files/metrics_by_network.rds")
null_networks = readRDS("../Data/Working_files/null_networks.rds")
null_metrics_networks = readRDS("../Data/Working_files/null_metrics_networks.rds")

# Prepare data
data = data %>%
dplyr::mutate(Year = lubridate::year(Date), 
              Month = lubridate::month(Date), 
              Day = lubridate::day(Date))

# Subset data for studies with only bumblebees
v = c("13_Karise", "22_Kallnik")
subsetting = unique(data$Study_id)

long_format = data %>% 
mutate(Network_id = paste0(Study_id, "_", Network_id, "_", Year)) %>% 
mutate(Network_id = str_replace_all(Network_id, " ","_")) %>% 
filter(!Study_id %in% v) %>% 
select(Network_id, Plant_accepted_name, Pollinator_accepted_name)

# Filter networks larger than 4 plants and 4 pollinator species
v2 = long_format %>% 
group_by(Network_id) %>% 
summarise(Poll_spp = n_distinct(Plant_accepted_name),
          Plant_spp = n_distinct(Pollinator_accepted_name)) %>% 
ungroup() %>% 
filter(Poll_spp >= 4 & Plant_spp >= 4) %>% 
pull(Network_id)

long_format = long_format %>% 
filter(Network_id %in% v2)

# Filter networks with more links than species
links = long_format %>% 
group_by(Network_id, Plant_accepted_name, Pollinator_accepted_name) %>%
distinct() %>% 
group_by(Network_id) %>% 
summarise(links = length(Network_id))

spp = long_format %>% 
group_by(Network_id) %>% 
summarise(n_plants = n_distinct(Plant_accepted_name),
          n_polls = n_distinct(Pollinator_accepted_name)) %>% 
mutate(n_total = n_plants + n_polls)

vector_to_include = left_join(links, spp)

v3 = vector_to_include %>% 
filter(links > n_total) %>% 
pull(Network_id)

long_format = long_format %>% 
filter(Network_id %in% v3)

# Explore how NODFc changes across the Y-X AXES (latitude-Longitude)
long_format_coords = data %>% 
mutate(Network_id = paste0(Study_id, "_", Network_id, "_", Year)) %>% 
mutate(Network_id = str_replace_all(Network_id, " ","_")) %>% 
select(Network_id, Latitude, Longitude, Bioregion) %>% 
distinct()

# Merge data
metrics_by_network_coords = left_join(metrics_by_network, long_format_coords)

# Connectance plots
# 1st Latitude
p1 = metrics_by_network_coords %>% 
ggplot(aes(Latitude, Connectance)) +
geom_point(aes(fill = Bioregion, shape = Bioregion), stroke = 0.15, size=2.5, alpha=1) +
theme_bw() +
ylab("Connectance") +
scale_fill_viridis_d() +
scale_colour_viridis_d() +
scale_shape_manual(values = c(
  "Alpine" = 21,  # Circle
  "Atlantic" = 22, # Triangle
  "Boreal" = 23,   # Square
  "Continental" = 24,  # Diamond
  "Mediterranean" = 25  # Cross
))+
ggtitle("(a)")

# 2nd Plot
p2 = metrics_by_network_coords %>% 
ggplot(aes(Latitude, Normalised_nestedness)) +
geom_point(aes(fill = Bioregion, shape = Bioregion), stroke = 0.15, size=2.5, alpha=1) +
theme_bw() +
ylab("NODFc") +
scale_fill_viridis_d() +
scale_colour_viridis_d() +
scale_shape_manual(values = c(
  "Alpine" = 21,  # Circle
  "Atlantic" = 22, # Triangle
  "Boreal" = 23,   # Square
  "Continental" = 24,  # Diamond
  "Mediterranean" = 25  # Cross
  )) +
ggtitle("(b)")

# Merge and process data
d = left_join(metrics_by_network, null_metrics_networks)
d = d %>% 
mutate(z_score = (Classic_nestedness - Mean_null_classic_nestedness) / Deviation_null_classic_nestedness)

p = 0.05 #cutoff probability 95% confidence
critical_value = qnorm(p/2) #double tail probability divide by 2

d1 = d %>%
mutate(infra_over_represented = case_when(
z_score < -abs(critical_value) ~ "Under-represented",
between(z_score, -abs(critical_value), abs(critical_value)) ~ "No statistical difference",
    z_score > abs(critical_value) ~ "Over-represented"
  ))

# 2nd Species (geometric mean)
species_number = long_format %>% 
group_by(Network_id) %>% 
summarise(N_plants = n_distinct(Plant_accepted_name),
          N_pollinators = n_distinct(Pollinator_accepted_name)) %>% 
group_by(Network_id) %>% 
summarise(geometric_mean_spp = geometric.mean(c(N_plants, N_pollinators)))

metrics_by_network_spp_number = left_join(metrics_by_network, species_number)

p3 = metrics_by_network_spp_number %>% 
ggplot(aes(geometric_mean_spp, Connectance)) +
geom_point(pch = 21, fill = "azure3", size=2.5, stroke=0.25, alpha=1) +
theme_bw() +
xlab("Species")+
ggtitle("(c)")

# Plot
p4 = d1 %>% 
ggplot(aes(x = z_score)) +
geom_histogram(aes(y = ..density.., fill = infra_over_represented), 
               bins = 100, alpha = 0.5, color = "black", position = "stack", size=0.1) +
stat_function(data = d1, fun = function(x) dnorm(x, mean = mean(d1$z_score), sd = sd(d1$z_score)) * 1.7, 
                n = 1000, inherit.aes = FALSE, color = "gray18", size = 1) +
theme_bw() +
coord_cartesian(expand = FALSE) +
geom_vline(xintercept = -abs(critical_value), linetype = "longdash", colour = "red")+
geom_vline(xintercept = abs(critical_value), linetype = "longdash", colour = "red") +
ylab("Density") +
xlab("Z scores (NODF)")  +
xlim(-6, 9) +
ylim(0, 0.8) +
scale_fill_manual(name = "Observed NODF against null",
                    limits = c("Under-represented", "No statistical difference", "Over-represented"),
                    labels = c("Less nested", "No difference", "More nested"),
                    values = c("Under-represented" = "coral2", 
                               "No statistical difference" = "palegreen3", 
                               "Over-represented" = "cyan3")) +
ggtitle("(d)")

# Layout plots
a = p1 + p2 + plot_layout(guides = 'collect') & theme(legend.position = "right",
                                                      legend.justification = "left")
b = (p3 + p4)  & theme(legend.justification = "left")
a / b  & theme(axis.title =  element_text(size=16, face="bold"),
               plot.title = element_text(size=22),
               panel.border = element_rect(colour = "black",
               fill=NA, size=1),
               axis.text=element_text(size=12),
               legend.text = element_text(size=10),
               legend.key.size = unit(6, "mm"),
               legend.spacing.y = unit(0.5, "cm"),
               legend.box.spacing = unit(0, "pt"),
               legend.title = element_text(face = "bold"))

@@

Although Europe...

Geographic and taxonomic biases!
Supporting reference:
https://doi.org/10.1111/ecog.05926

Networks, ref:

https://www.biorxiv.org/content/10.1101/2020.04.02.021691v1.abstract

Major changes in NW Europe occurred pre-1990
https://doi.org/10.1111/ele.12121

EuPPollNet represents the largest database of plant-pollinator interactions compiled to date and is a work in progress database as there are more studies, unsampled habitats, species, etc...

Contextualise European diversity of plant and pollinator species with global biodiveristy

>>>>

The interaction between plants and pollinators is one of the most well-documented mutualisms on Earth. Plant-pollinator interactions involve a great diversity of species, largely attributed to their coevolutionary history [@ollerton2017], and are critically important for terrestrial biodiversity and the human economy. The synergistic effects of climate change with other global change pressures are threatening worldwide biodiversity [@sala2000; @bellard2014], including plant and pollinator species as well as their interactions [@goulson2015; @settele2016; @eichenberg2021]. Under this scenario, the increasing availability of biodiversity data plays a major role in our ecological understanding of species and guiding conservation planning [@heberling2021]. However, our knowledge of plant and pollinator species and their network of interactions still exhibits major temporal, spatial and taxonomic biases [@archer2014; @troia2016; @poisot2021; @marshall2024], limiting our ability to effectively protect their biodiversity.

@@

Here, we present the EuPPollNet database, an acronym derived from European plant-pollinator networks, which contains harmonized information on interaction data of plants and pollinators at European level. The primary focus of the pollinator taxonomic groups is on the main orders of entomofauna that visit and pollinate flowering plants in Europe. These include insect species from the orders Hymenoptera, Diptera, Lepidoptera, and Coleoptera [@willemstein1987; @potts2015], which comprise almost the totality of recorded interactions in the EuPPollNet database (`r format(round(main_orders_percent, 2), nsmall = 2)`%). To understand the scope of the database, we examine the taxonomic and sampling coverage of the different plant and pollinator species at European level with the help of the most up-to-date species checklists and rarefaction analyses. In addition, we investigate how two key structural metrics of plant-pollinator networks, such as connectance and nestedness, change across the latitudinal range of studies and bioclimatic regions. Finally, we also examine whether plant-pollinator networks fulfill the expectation of being more nested than null expectations. EuPPollNet aims to cover a wide range of taxonomic groups and habitats, while also providing other variables of interest that allow better control for the ecological context and sampling methods. In addition, EuPPollNet offers a transparent and accessible workflow of its data management and species harmonization that allows it to be reused and keep building on it over time. This database provides a large number of community-level networks with curated and harmonized data, distinguishing it from other currently available resources that contain plant-pollinator interactions. We expect that this database will help to evaluate macro-ecological processes and current gaps of plant-pollinator interactions at European level.

@@

For plants: (i) we initially verified the exact matches against the GBIF species checklist; (ii) selected unmatched cases and fix orthographic errors; (iii) retrieved again taxonomic information for those unmatched cases, evaluated accuracy of fuzzing matching and programmatically fixed records that are still not found; (iv) finally, we used the World Flora Taxonomic Backbone [@govaerts2021\; WFO, July 7, 2022] as the ultimate filter for taxonomic information as we used it to calculate the plant taxonomic coverage of our database.

For pollinators: (i) we first created a checklist of species names for the most representative pollinator groups at the European level by combining the most up to date published checklists of bees and syrphids [@reverte2023] with the one of butterflies [@wiemers2018]; (ii) then, we compared pollinator species names against the checklist and recovered some unmatched cases with restrictive fuzzy matching by using **stringdist** package [@van2014]; (iii) we programmatically fixed unmatched records when necessary and retrieved the taxonomic information for all species from GBIF; (iv) we fixed the non-found cases in the GBIF checklist and made sure that all species names from bees, syrphids and butterflies were named according to their respective species checklists.

@@

To assess the completeness of plant and pollinator species in the EuPPollNet database at European level, we used the aforementioned checklists for plants and pollinators. Specifically for plants, we refined the checklist to include only European flowering plants and excluded taxonomic groups not associated with biotic pollination. We did this by first excluding the families considered to have exclusively a wind pollination mode [see @culley2002], and then by filtering out the genera with wind or non-biotic pollination from families that exhibit both biotic and non-biotic pollination modes. Additionally, we manually included some exotic species and added unresolved species names that were not present in the accepted names of the checklist at the current version of usage. For pollinators, we compared only the taxonomic coverage of bees, syrphids and butterflies by using their species checklists at European level. The potential number of pollinator species at European level was estimated by adding the total number of species of bees, syrphids, and flies from the checklists, along with the extrapolated number of species from other insect pollinators. This extrapolated number was estimated by assuming that the sampling coverage of these other insect pollinators is equal to the average coverage across bees, syrphids, and butterflies (mean coverage = `r round(mean_coverage, 2)`; sd = `r round(sd_coverage,2)`).

Finally, to evaluate if the presence-absence of interaction records for bees and flowering plants follows a phylogenetic pattern within the database, we calculated its phylogenetic signal  at genus and family level, respectively. The phylogenetic signal was calculated by using the _phylosig_ function from the **phytools** package [@revell2012]. We extracted the phylogenetic information for bees from a genus level phylogeny [@hedtke2013] and processed it with the help of the packages **ape** [@paradis2019], **MCMCglmm** [@hadfield2010] and **phytools**; and for plants the phylogenetic tree was obtained from a species level plant phylogeny [@smith2018] with the help of **rtree** package [@li2023].

@@

The estimated sampling coverage of plant and pollinator species within the database across the different sampling sites is approximately `r round(poll_coverage)`% for both taxonomic groups. This suggests that the rarefied accumulation curves of both plant and pollinator species exhibit already a "quasi-asymptotic" growth of species richness by considering the current number of sampling sites or networks (**Figure 2a-2b**). The predicted observed species richness by doubling the sampling effort on the already sampled habitat types within the database will only increase pollinator richness by `r format(round(poll_increase3, 2), nsmall=2)`% and plant richness by `r format(round(plant_increase3, 2), nsmall=2)`%. However, the sampling coverage of interactions is `r format(round(int_coverage, 2), nsmall=2)`%, and by doubling the sampling effort the predicted number of unique interactions recorded will have approximately a twofold increase (`r format(round(int_increase3, 2), nsmall=2)`%; **Figure 2c**). Similarly, to the plant and pollinator species accumulation curves across sampling sites, when we consider the accumulated pollinator richness across sampled plant species, this curve also shows a "quasi-asymptotic" growth with a sampling coverage value of `r format(round(plant_poll_coverage, 2), nsmall=2)`%. The predicted recorded pollinator species by doubling the number of plants sampled is expected to increase by `r format(round(plant_poll_increase3, 2), nsmall=2)`% (**Figure 2d**). In addition, when we explore how common or rare are plant and pollinator species across all sampling sites, we find that the majority of plant and pollinator species tend to be regionally specific, and only a minor portion of them is being shared across a broad range of sampling sites (**Figures 2e-2f**). For instance, most plants (`r format(round(plants_under_one_percent, 2), nsmall=2)`%) and pollinators (`r format(round(polls_under_one_percent, 2), nsmall=2)`%) are exclusively found in less than 1% of sampling sites; and the most common plant (*`r most_common_plant`*) and pollinator (*`r most_common_poll`*) species are only found in `r format(round(most_common_plant_percent, 2), nsmall=2)`% and `r format(round(most_common_poll_percent, 2), nsmall=2)`% of sampling sites, respectively.

```{r echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE, out.width="95%", fig.cap= "\\textbf{Figure 2}. Graphs (a-b-c) indicate the accumulation curves for pollinators, plants, and their interactions across sampling sites. Grey solid lines represent 100 randomized accumulation curves, the black solid lines represent the interpolated curve (i.e., the mean across curves), and the red dashed lines illustrate the extrapolated curve for approximately 3000 sampling sites. The solid black points indicate the number of species and interactions contained in the database. Graph (d) shows the accumulation curve of pollinator species across an increasing number of plant species. This last graph uses the same color and shape structure as the ones in the top panel. Graphs (e-f) indicate the percentage of occurrence (i.e., incidence) of plant and pollinator species across sampling sites. Species on the left (i.e., common) are found in many sampling sites, while species on the right (i.e., rare) are found in few or only a single sampling site. Note that indeed \\textit{Apis mellifera} is the most common pollinator but was excluded from this visualization.", fig.width=10, fig.height=6, out.width="115%"}

@@

Connectance values ranged between `r round(min_connectance, 2)` to `r round(max_connectance, 2)` ($\bar{x}$ = `r round(mean_connectance, 2)`) and as expected, followed an negative exponential relationship with the number of species per network (Kendall $\tau$ = `r round(connectance_spp_mean_corr_tau, 2)`, _P_ < 0.01; **Figure 6a**). Nestedness values (NODFc) ranged between `r round(min_nodfc, 2)` to `r round(max_nodfc, 2)` ($\bar{x}$ = `r round(mean_nodfc, 2)`), and as expected for this metric, were independent of the mean number of species (Kendall $\tau$ = `r round(normalised_nestedness_spp_mean_corr_tau, 2)`, _P_ = `r round(normalised_nestedness_spp_mean_corr_pval, 2)`; **Figure S3**). We found that only `r round(over_z_percent, 2)`% of networks were statistically more nested than expected by chance, with `r round(no_diff_z_percent, 2)`% showing no statistical difference, and none being less nested than null expectations (**Figure 6b**). Latitude explained little of the observed variability of residual connectance and nestedness across networks (connectance: R$^2$ = `r round(connectance_lat_r2, 2)`, _P_ = `r round(connectance_lat_pval, 2)`, **Figure 6a**; NODFc: R$^2$ = `r round(normalised_nestedness_lat_r2, 2)`, _P_ = `r round(normalised_nestedness_lat_pval, 2)`, **Figure 6b**). Overall, networks towards higher latitudes showed lower residual connectance and higher nestedness that networks located in lower latitudes. Note that residual connectance and normalised nestedness showed a moderate negative correlation (Kendall $\tau$ = `r round(resid_connectance_nestedness_corr_tau, 2)`, _P_ = `r round(resid_connectance_nestedness_corr_pval, 2)`).

```{r echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE, out.width="95%", fig.cap= "\\textbf{Figure 6}. Graph (a) shows the association between network connectance and the geometric mean of plant and pollinator species per network on a log-scale with the respective fitted line from a Beta regression. Graph (b) shows the distribution of z-scores when comparing the nestedness from the empirical networks with their randomised counterparts (100 null models for each network). The vertical red dashed lines represent the z critical value for a two tailed test with alpha = 0.05. Z-scores to the left of the first vertical red dashed line indicate that networks are less nested than expected by chance (red), those between the two dashed lines indicate no statistical difference from random expectations (green), and those to the right indicate that networks are more nested than expected by chance (blue). Graphs (c-d) show the fitted regression of residual connectance and nestedness across the latitudinal range of the studies from the database. In addition, the bioclimatic region of each network is indicated with points of different shapes and colours.", fig.width=10, fig.height=7, fig.align='center'}

# Load libraries
library(dplyr)
library(ggplot2)
library(stringr)
library(patchwork)

# Load plots generated in scripts 7_2_Nestedness and 7_2_Connectance

connectance_species = readRDS("../Data/Manuscript_info/Figure6_connectance_species.rds")
connectance_latitude = readRDS("../Data/Manuscript_info/Figure6_connectance_latitude.rds")
nestedness_z_scores = readRDS("../Data/Manuscript_info/nestedness_z_scores.rds")
nestedness_latitude = readRDS("../Data/Manuscript_info/nestedness_latitude.rds")

#Prepare first connectance plots
library(betareg)
model = betareg(Connectance ~ log_geometric_mean_spp, data = connectance_species)
p1 = ggplot(connectance_species, aes(x = log_geometric_mean_spp, y = Connectance))+
geom_point(pch = 21, fill = "azure3", size=2.5) +
geom_line(aes(y = predict(model, connectance_species)), colour= "black")+
xlab("log(Species)") +
theme_bw() +
ggtitle("(a)")

p3 = connectance_latitude %>% 
ggplot(aes(Latitude, residual_conectance)) +
geom_point(aes(fill = Bioregion, shape = Bioregion), stroke = 0.25, size=2.5) +
scale_fill_viridis_d() +
scale_colour_viridis_d() +
scale_shape_manual(values = c(
  "Alpine" = 21,  # Circle
  "Atlantic" = 22, # Triangle
  "Boreal" = 23,   # Square
  "Continental" = 24,  # Diamond
  "Mediterranean" = 25  # Cross
  )) +
geom_line(data = connectance_latitude,aes(x=Latitude, y = fit), inherit.aes = FALSE)+
geom_ribbon(data = connectance_latitude, 
    aes(x=Latitude,ymin = lwr, ymax = upr), inherit.aes = FALSE,
    fill = "grey70", alpha= 0.3) +
theme_bw()+
ylab("Residual connectance") +
ggtitle("(c)")

#Prepare now nestedness plots
# Plot
p = 0.05 #cutoff probability 95% confidence
critical_value = qnorm(p/2) #double tail probability divide by 2

nestedness_z_scores = nestedness_z_scores %>%
mutate(infra_over_represented = case_when(
z_score < -abs(critical_value) ~ "Under-represented",
between(z_score, -abs(critical_value), abs(critical_value)) ~ "No statistical difference",
    z_score > abs(critical_value) ~ "Over-represented"
  ))

p2 = nestedness_z_scores %>% 
ggplot(aes(x = z_score)) +
geom_histogram(aes(y = ..density.., fill = infra_over_represented), 
               bins = 100, alpha = 0.5, color = "black", position = "stack") +
stat_function(data = nestedness_z_scores, fun = function(x) dnorm(x, mean = mean(nestedness_z_scores$z_score), sd = sd(nestedness_z_scores$z_score)) * 1.7, 
              n = 1000, inherit.aes = FALSE, color = "gray18", size = 1) +
theme_bw() +
coord_cartesian(expand = FALSE) +
geom_vline(xintercept = -abs(critical_value), linetype = "longdash", colour = "red") +
geom_vline(xintercept = abs(critical_value), linetype = "longdash", colour = "red") +
ylab("Density") +
xlab("Z scores (NODF)")  +
xlim(-6, 9) +
ylim(0, 0.8) +
scale_fill_manual(name = "Observed against null",
                  limits = c("Under-represented", "No statistical difference", "Over-represented"),
                  labels = c("Less nested", "No difference", "More nested"),
                  values = c("Under-represented" = "coral2", 
                             "No statistical difference" = "palegreen3", 
                             "Over-represented" = "cyan3")) +
ggtitle("(b)")

# 2nd Plot
p4 = nestedness_latitude %>% 
ggplot(aes(Latitude, Normalised_nestedness)) +
geom_point(aes(fill = Bioregion, shape = Bioregion), stroke = 0.25, size=2.5) +
theme_bw() +
ylab("NODFc") +
scale_fill_viridis_d() +
scale_colour_viridis_d() +
scale_shape_manual(values = c(
  "Alpine" = 21,  # Circle
  "Atlantic" = 22, # Triangle
  "Boreal" = 23,   # Square
  "Continental" = 24,  # Diamond
  "Mediterranean" = 25  # Cross
  )) +
geom_line(data = nestedness_latitude,aes(x=Latitude, y = fit), inherit.aes = FALSE)+
geom_ribbon(data = nestedness_latitude, 
    aes(x=Latitude,ymin = lwr, ymax = upr), inherit.aes = FALSE,
    fill = "grey70", alpha= 0.3) +
theme_bw()+
ggtitle("(d)")

# Layout plots
#a = (p1 + p2) 
#b = p3 + p4 + plot_layout(guides = 'collect') & theme(legend.position = "right")
#a / b

# Layout plots
a = (p1 + p2)  & theme(legend.position = "right", legend.justification = "left")
b = p3 + p4 + plot_layout(guides = 'collect') & theme(legend.justification = "left")
a / b  & theme(axis.title =  element_text(size=16, face="bold"),
               plot.title = element_text(size=22),
               panel.border = element_rect(colour = "black",
               fill=NA, size=1),
               axis.text=element_text(size=12))

@@

EuPPollNet offers the largest set of plant-pollinator networks and plant-pollinator studies compiled to date at European level. The database contains `r prettyNum(plant_spp_number, big.mark=",")` plant and `r prettyNum(pollinator_spp_number, big.mark=",")` pollinator species with over a million of interaction records. While the overall sampling coverage of species and interactions is relatively high across the sampled sites, the taxonomic coverage of plants and the main pollinator groups at the European level is still relatively low (i.e., `r round(plant_coverage, 1)`% for flowering plants and `r round(bee_coverage, 1)`% for bee species). Given that most of the plant-pollinator networks from the database are sampled on pastures, and habitat heterogeneity is a crucial factor in understanding pollinator diversity at European level [@kleijn2015; @hass2018; @martinez2022], adding studies on other habitat types is likely to result in a rapid increase of the coverage of plant and pollinator species and their interactions. However, plant and pollinator species were rarely shared across multiple sites, indicating that there are few "common" species and many "rare" ones at the metaweb or continental level. This high number of "rare" species results in an upward slope of the species or interaction accumulation curves [@thompson2003]. In other words, minimal sampling efforts are capturing a substantial number of species and interactions, but achieving a comprehensive inventory will require of numerous sampling events within and across habitats, particularly for plant-pollinator interactions.

Bees conducted the majority of the interactions at the metaweb level, but their relative relevance changed across habitats and bioclimatic regions. For instance, plant-pollinator communities in the Mediterranean were dominated by bees, while communities in Alpine or Boreal regions were especially fly species rich or fly-dominated. These patterns are consistent with our current understanding of bee diversity, which peaks in dry or temperate areas [@orr2021; @leclercq2023]; and with the fact that colder environments (i.e., altitude and latitude wise) harbor a larger fraction of fly pollinators compared to other taxa [@elberling1999; @lefebvre2018]. In addition, beetles were only species rich in the Mediterranean region. Although the networks from the database are visitation networks and do not capture pollinator efficiency [@ballantyne2015], the high proportion of beetles as floral visitors in the Mediterranean provides further support for their role as pollinators in this region [@herrera2019; @leon2022]. The number of butterfly species and interactions were relatively low compared to the other taxa. While Europe contains fewer butterfly species than other regions of the world [@ollerton2017], their relevance as pollinators is likely underestimated within this database. This is because a large fraction of studies (~40%) did not sample butterflies, and conventional sampling methods for monitoring other insect pollinators (e.g., bees or flies) may be inadequate for sampling plant-butterfly interactions [@isaac2011]. Honey bees were present in `r round(apis_percent_networks, 1)`% of networks and conducted in average a third of the total interactions per network. Indeed, this proportion is higher that the one found when considering only natural communities at a global scale [~13%\; @hung2018], which highlights their key role as potential pollinator of many flowering plants. However, high honey bee densities can have negative impacts on wild pollinators and the plants they visit [@steffan2000; @geldmann2018; @herrera2020], emphasising the need to evaluate these impacts and implement actions if necessary.
